---
title: "Analysing the Waitlist and Decline Groups"
format:
  html:
    code-fold: true
    embed-resources: true
    warning: false
    echo: false
---

```{r}
library(tidyverse)
library(corrplot)
library(gt)
```

## Overview of Waitlist vs Decline Groups

After exploring some of the trends between students who were admitted versus students who were declined or waitlisted, we felt it was also important to look more at what the difference was between a student getting declined or waitlisted. First, here are the differences between the averages of all of the testing scores, GPA, volunteer levels, and work experience between these two groups. 

```{r}
data <- read_csv("data/cleaned_data/data_cleaned.csv")
wd <- data[data$Decision != "Admit", ]
num_vars <- subset(wd, select = -c(State, ...1))
num_vars <- subset(wd, select = -c(State, ...1))
num_vars %>% group_by(Decision) %>%
  summarise("GPA" = mean(GPA), 
            "Work Experience" = mean(WorkExp), 
            "Test Score" = mean(TestScore), 
            "Writing Score" = mean(WritingScore), 
            "Volunteer Level" = mean(VolunteerLevel)) %>% 
  gt() %>%
  tab_style(
    style = cell_fill(color = "#FFF3CD"),  
    locations = cells_body(
      columns = GPA,
      rows = GPA == max(GPA)
    )
  ) %>%
  tab_style(
    style = cell_fill(color = "#FFF3CD"),
    locations = cells_body(
      columns = `Test Score`,
      rows = `Test Score` == max(`Test Score`)
    )
  ) %>%
  tab_style(
    style = cell_fill(color = "#FFF3CD"),
    locations = cells_body(
      columns = `Work Experience`,
      rows = `Work Experience` == max(`Work Experience`)
    ) 
  ) %>%
  tab_style(
    style = cell_fill(color = "#FFF3CD"),
    locations = cells_body(
      columns = `Volunteer Level`,
      rows = `Volunteer Level` == max(`Volunteer Level`)
    )
  ) %>% 
    tab_style(
    style = cell_fill(color = "#FFF3CD"),
    locations = cells_body(
      columns = `Writing Score`,
      rows = `Writing Score` == max(`Writing Score`)
    )
  )
```

From this table, we can see the students on the waitlist have higher scores in each categories. There appear to be bigger differences between GPA, Test Scores, and Writing Scores, than in work experience and volunteer level. We can do a statistical test to test for a difference in means. 

```{r}
gpa_tt <- t.test(GPA ~ Decision, data = wd, var.equal = FALSE) 
work_tt <- t.test(WorkExp ~ Decision, data = wd, var.equal = FALSE)
test_tt <- t.test(TestScore ~ Decision, data = wd, var.equal = FALSE)
writ_tt <- t.test(WritingScore ~ Decision, data = wd, var.equal = FALSE)
volunteer_tt <- t.test(VolunteerLevel ~ Decision, data = wd, var.equal = FALSE)

tests <- c("GPA", "WorkExp", "TestScore", "WritingScore", "VolunteerLevel")
p_vals <- data.frame(
  variable = character(),    
  p_value = numeric(),       
  stringsAsFactors = FALSE   
)

for (variable in tests) {
  formula <- as.formula(paste(variable, "~ Decision"))
  test <- t.test(formula, data = wd, var.equal = FALSE) 
  new_row <- data.frame(
    variable = variable,
    p_value = test$p.value,
    p_value_formatted = "",  
    stringsAsFactors = FALSE
  )
  
  if (new_row$p_value < 0.001) {
    new_row$p_value_formatted <- "< 0.001"
  } else {
    new_row$p_value_formatted <- sprintf("%.3f", new_row$p_value)
  }
  
  p_vals <- rbind(p_vals, new_row)
}

p_vals %>%
  select(variable, p_value_formatted) %>%  
  rename("Variable" = variable,
         "p-value" = p_value_formatted) %>%
  gt() %>%
  tab_style(
    style = cell_fill(color = "#FFF3CD"), 
    locations = cells_body(
      columns = "p-value",
      rows = p_vals$p_value < 0.05  
    )
  )

```

## Exploring Significant Variables 

```{r}
wd_bands <- data.frame(wd)
wd_bands <- subset(wd_bands, select = -c(...1))
wd_bands <- wd_bands %>% 
  mutate(GPA_Bands = case_when(
    GPA < 2.40 ~ "< 2.40", 
    GPA >= 2.40 & GPA <= 2.49 ~ "2.40-2.49", 
    GPA >= 2.50 & GPA <= 2.59 ~ "2.50-2.59", 
    GPA >= 2.60 & GPA <= 2.69 ~ "2.60-2.69", 
    GPA >= 2.70 & GPA <= 2.79 ~ "2.70-2.79", 
    GPA >= 2.80 & GPA <= 2.89 ~ "2.80-2.89", 
    GPA >= 2.90 & GPA <= 2.99 ~ "2.90-2.99", 
    GPA >= 3.00 & GPA <= 3.09 ~ "3.00-3.09", 
    GPA >= 3.10 & GPA <= 3.19 ~ "3.10-3.19", 
    GPA >= 3.20 & GPA <= 3.29 ~ "3.20-3.29", 
    GPA >= 3.30 & GPA <= 3.39 ~ "3.30-3.39", 
    GPA >= 3.40 & GPA <= 3.49 ~ "3.40-3.49", 
    GPA >= 3.50 & GPA <= 3.59 ~ "3.50-3.59", 
    GPA >= 3.60 & GPA <= 3.69 ~ "3.60-3.69", 
    GPA >= 3.70 & GPA <= 3.79 ~ "3.70-3.79", 
    GPA >= 3.80 & GPA <= 3.89 ~ "3.80-3.89", 
    GPA >= 3.90 & GPA <= 3.99 ~ "3.90-3.99", 
    GPA == 4.00 ~ "4.00"
  ))

wd_bands_count <- wd_bands %>%
  count(GPA_Bands, Decision) %>%
  complete(GPA_Bands, Decision, fill = list(n = 0))

ggplot(wd_bands_count, aes(fill = Decision, y = n, x = GPA_Bands)) + 
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("Decline" = "firebrick", "Waitlist" = "darkgoldenrod1")) +
  theme_minimal() +
  labs(x = "GPA Bands", y = "Count", fill = "Decision") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ggtitle("Comparison of Waitlist vs Decline GPA")
```

```{r}
wd_bands <- data.frame(wd)
wd_bands <- subset(wd_bands, select = -c(...1))
wd_bands <- wd_bands %>% 
  mutate(Test_Score_Bands = case_when(
    TestScore < 760 ~ "< 760", 
    TestScore >= 760 & TestScore <= 769 ~ "760-769", 
    TestScore >= 770 & TestScore <= 779 ~ "770-779", 
    TestScore >= 780 & TestScore <= 789 ~ "780-789", 
    TestScore >= 790 & TestScore <= 799 ~ "790-799", 
    TestScore >= 800 & TestScore <= 809 ~ "800-809", 
    TestScore >= 810 & TestScore <= 819 ~ "810-819", 
    TestScore >= 820 & TestScore <= 829 ~ "820-829", 
    TestScore >= 830 & TestScore <= 839 ~ "830-839", 
    TestScore >= 840 & TestScore <= 849 ~ "840-849", 
    TestScore >= 850 & TestScore <= 859 ~ "850-859", 
    TestScore >= 860 & TestScore <= 869 ~ "860-869", 
    TestScore >= 870 & TestScore <= 879 ~ "870-879", 
    TestScore >= 880 & TestScore <= 889 ~ "880-889", 
    TestScore >= 890 & TestScore <= 899 ~ "890-899", 
    TestScore > 900 ~ "> 900"
  ))


wd_bands_count <- wd_bands %>%
  count(Test_Score_Bands, Decision) %>%
  complete(Test_Score_Bands, Decision, fill = list(n = 0))

ggplot(wd_bands_count, aes(fill = Decision, y = n, x = Test_Score_Bands)) + 
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("Decline" = "firebrick", "Waitlist" = "darkgoldenrod1")) +
  theme_minimal() +
  labs(x = "Test Score Bands", y = "Count", fill = "Decision") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ggtitle("Comparison of Waitlist vs Decline Test Scores")
```

```{r}
wd_bands <- data.frame(wd)
wd_bands <- subset(wd_bands, select = -c(...1))
wd_bands <- wd_bands %>% 
  mutate(Writing_Score_Bands = case_when(
    WritingScore < 60 ~ "< 60", 
    WritingScore >= 60 & WritingScore <= 64 ~ "60-64", 
    WritingScore >= 65 & WritingScore <= 69 ~ "65-69", 
    WritingScore >= 70 & WritingScore <= 74 ~ "70-74", 
    WritingScore >= 75 & WritingScore <= 79 ~ "75-79", 
    WritingScore >= 80 & WritingScore <= 84 ~ "80-84", 
    WritingScore >= 85 & WritingScore <= 89 ~ "85-89", 
    WritingScore > 90 ~ "> 90"
  ))

wd_bands_count <- wd_bands %>%
  count(Writing_Score_Bands, Decision) %>%
  complete(Writing_Score_Bands, Decision, fill = list(n = 0))

ggplot(wd_bands_count, aes(fill = Decision, y = n, x = Writing_Score_Bands)) + 
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(values = c("Decline" = "firebrick", "Waitlist" = "darkgoldenrod1")) +
  theme_minimal() +
  labs(x = "Writing Score Bands", y = "Count", fill = "Decision") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ggtitle("Comparison of Waitlist vs Decline Writing Scores")
```


## Case Study

```{r}
decline <- wd[wd$Decision == "Decline", ]
decline <- subset(decline, select = -c(...1))
max_gpa_d <- decline[decline$GPA == max(decline$GPA), ] 
max_gpa_d %>% gt()
```



```{r}
admit <- data[data$Decision == "Admit", ]
admit <- subset(admit, select = -c(...1))
max_gpa_a <- admit[admit$GPA >= 3.80 & admit$GPA <= 3.90, ][2, ]
max_gpa_a %>% gt()
```

```{r}
waitlist <- data[data$Decision == "Waitlist", ]
waitlist <- subset(waitlist, select = -c(...1))
max_gpa_w <- waitlist[waitlist$GPA == max(waitlist$GPA), ]
max_gpa_w %>% gt()
```


Based on GPA, we may be surprised that candidate 1 got declined. As we have seen, a 3.84 is well within the range for acceptance. Their test scores and writing scores were also on the higher end of the waitlist range. The question remains, why did this person get declined and not admitted, let alone waitlisted? The main difference we see between this candidate and the admitted candidates are the test and writing scores. Were these scores higher, perhaps they would have been admitted. There also was a difference in volunteer level. 



